---
title: "Defining mutational signatures of lung cancer-associated carcinogens through in vitro exposure of human airway epithelial cells"
author: "Natasha Gurevich"
date: "2026-02-24"
output: word_document
---

# Setup

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(musicatk)
library(plyr)
library(factoextra)
library(ggrepel)
library(pca.utils)
library(ggpubr)
library(corrplot)
library(RColorBrewer)
library(lsa)
library(rstatix)
library(tidyverse)
library(readxl)

```

# Load Variants

```{r}

#establish file paths
C11_vcf_file_path <- "../data/vcf/final_rerun/C1-1-vs-C1-3-filtered.vcf"
C12_vcf_file_path <- "../data/vcf/final_rerun/C1-2-vs-C1-3-filtered.vcf"
C13_vcf_file_path <- "../data/vcf/final_rerun/C1-3-vs-C1-2-filtered.vcf"
C21_vcf_file_path <- "../data/vcf/final_rerun/C2-1-vs-C1-3-filtered.vcf"
C22_vcf_file_path <- "../data/vcf/final_rerun/C2-2-vs-C1-3-filtered.vcf"
C23_vcf_file_path <- "../data/vcf/final_rerun/C2-3-vs-C1-3-filtered.vcf"
C31_vcf_file_path <- "../data/vcf/final_rerun/C3-1-vs-C1-3-filtered.vcf"
C32_vcf_file_path <- "../data/vcf/final_rerun/C3-2-vs-C1-3-filtered.vcf"
C33_vcf_file_path <- "../data/vcf/final_rerun/C3-3-vs-C1-3-filtered.vcf"
C61_vcf_file_path <- "../data/vcf/final_rerun/C6-1-vs-C2-1-filtered.vcf"
C62_vcf_file_path <- "../data/vcf/final_rerun/C6-2-vs-C2-1-filtered.vcf"
C63_vcf_file_path <- "../data/vcf/final_rerun/C6-3-vs-C2-1-filtered.vcf"
C64_vcf_file_path <- "../data/vcf/final_rerun/C6-4-vs-C2-1-filtered.vcf"
C71_vcf_file_path <- "../data/vcf/final_rerun/C7-1-vs-C3-1-filtered.vcf"
C72_vcf_file_path <- "../data/vcf/final_rerun/C7-2-vs-C3-1-filtered.vcf"
C73_vcf_file_path <- "../data/vcf/final_rerun/C7-3-vs-C3-1-filtered.vcf"
C74_vcf_file_path <- "../data/vcf/final_rerun/C7-4-vs-C3-1-filtered.vcf"
C81_vcf_file_path <- "../data/vcf/final_rerun/C8-1-vs-C2-1-filtered.vcf"
C82_vcf_file_path <- "../data/vcf/final_rerun/C8-2-vs-C2-1-filtered.vcf"
C83_vcf_file_path <- "../data/vcf/final_rerun/C8-3-vs-C2-1-filtered.vcf"
C84_vcf_file_path <- "../data/vcf/final_rerun/C8-4-vs-C2-1-filtered.vcf"
C91_vcf_file_path <- "../data/vcf/final_rerun/C9-1-vs-C2-1-filtered.vcf"
C92_vcf_file_path <- "../data/vcf/final_rerun/C9-2-vs-C2-1-filtered.vcf"
C93_vcf_file_path <- "../data/vcf/final_rerun/C9-3-vs-C2-1-filtered.vcf"
C94_vcf_file_path <- "../data/vcf/final_rerun/C9-4-vs-C2-1-filtered.vcf"

# read in variants
C11_variants <- extract_variants_from_vcf_file(vcf_file = C11_vcf_file_path, id = "C1-1", rename = "control_rep1")
C12_variants <- extract_variants_from_vcf_file(vcf_file = C12_vcf_file_path, id = "C1-2", rename = "control_rep2")
C13_variants <- extract_variants_from_vcf_file(vcf_file = C13_vcf_file_path, id = "C1-3", rename = "control_rep3")
C21_variants <- extract_variants_from_vcf_file(vcf_file = C21_vcf_file_path, id = "C2-1", rename = "control_DMSO_rep1")
C22_variants <- extract_variants_from_vcf_file(vcf_file = C22_vcf_file_path, id = "C2-2", rename = "control_DMSO_rep2")
C23_variants <- extract_variants_from_vcf_file(vcf_file = C23_vcf_file_path, id = "C2-3", rename = "control_DMSO_rep3")
C31_variants <- extract_variants_from_vcf_file(vcf_file = C31_vcf_file_path, id = "C3-1", rename = "control_acetone_rep1")
C32_variants <- extract_variants_from_vcf_file(vcf_file = C32_vcf_file_path, id = "C3-2", rename = "control_acetone_rep2")
C33_variants <- extract_variants_from_vcf_file(vcf_file = C33_vcf_file_path, id = "C3-3", rename = "control_acetone_rep3")
C61_variants <- extract_variants_from_vcf_file(vcf_file = C61_vcf_file_path, id = "C6-1", rename = "BaP_5nM_DMSO_rep1")
C62_variants <- extract_variants_from_vcf_file(vcf_file = C62_vcf_file_path, id = "C6-2", rename = "BaP_5nM_DMSO_rep2")
C63_variants <- extract_variants_from_vcf_file(vcf_file = C63_vcf_file_path, id = "C6-3", rename = "BaP_5nM_DMSO_rep3")
C64_variants <- extract_variants_from_vcf_file(vcf_file = C64_vcf_file_path, id = "C6-4", rename = "BaP_5nM_DMSO_rep4")
C71_variants <- extract_variants_from_vcf_file(vcf_file = C71_vcf_file_path, id = "C7-1", rename = "NTCU_0.5nM_Acetone_rep1")
C72_variants <- extract_variants_from_vcf_file(vcf_file = C72_vcf_file_path, id = "C7-2", rename = "NTCU_0.5nM_Acetone_rep2")
C73_variants <- extract_variants_from_vcf_file(vcf_file = C73_vcf_file_path, id = "C7-3", rename = "NTCU_0.5nM_Acetone_rep3")
C74_variants <- extract_variants_from_vcf_file(vcf_file = C74_vcf_file_path, id = "C7-4", rename = "NTCU_0.5nM_Acetone_rep4")
C81_variants <- extract_variants_from_vcf_file(vcf_file = C81_vcf_file_path, id = "C8-1", rename = "NNK_0.01nM_DMSO_rep1")
C82_variants <- extract_variants_from_vcf_file(vcf_file = C82_vcf_file_path, id = "C8-2", rename = "NNK_0.01nM_DMSO_rep2")
C83_variants <- extract_variants_from_vcf_file(vcf_file = C83_vcf_file_path, id = "C8-3", rename = "NNK_0.01nM_DMSO_rep3")
C84_variants <- extract_variants_from_vcf_file(vcf_file = C84_vcf_file_path, id = "C8-4", rename = "NNK_0.01nM_DMSO_rep4")
C91_variants <- extract_variants_from_vcf_file(vcf_file = C91_vcf_file_path, id = "C9-1", rename = "NNK_1.0nM_DMSO_rep1")
C92_variants <- extract_variants_from_vcf_file(vcf_file = C92_vcf_file_path, id = "C9-2", rename = "NNK_1.0nM_DMSO_rep2")
C93_variants <- extract_variants_from_vcf_file(vcf_file = C93_vcf_file_path, id = "C9-3", rename = "NNK_1.0nM_DMSO_rep3")
C94_variants <- extract_variants_from_vcf_file(vcf_file = C94_vcf_file_path, id = "C9-4", rename = "NNK_1.0nM_DMSO_rep4")

# aggreagate variants from all samples
all_variants <- rbind(C11_variants, C12_variants, C13_variants,
                      C21_variants, C22_variants, C23_variants,
                      C31_variants, C32_variants, C33_variants,
                      C61_variants, C62_variants, C63_variants, C64_variants,
                      C71_variants, C72_variants, C73_variants, C74_variants,
                      C81_variants, C82_variants, C83_variants, C84_variants,
                      C91_variants, C92_variants, C93_variants, C94_variants)

# select genome
g <- select_genome("hg38")

# create musica object
musica_all <- create_musica_from_variants(x = all_variants, genome = g)

# build mutation count tables
build_standard_table(musica_all, g = g, modality = "SBS96")
build_standard_table(musica_all, g = g, modality = "DBS78")
build_standard_table(musica_all, g = g, modality = "IND83")

# add annotations to group by
samp_annot(musica_all, "Exposure") <- c(rep("BaP", 4), rep("control_acetone", 3), rep("control_DMSO", 3), rep("control", 3), rep("NNK_0.01", 4), rep("NNK_1.0", 4), rep("NTCU", 4))


```

# K value assist to determine number of signatures

```{r}

sbs_k_trial <- compare_k_vals(musica_all, modality = "SBS96", reps = 50, min_k = 1, max_k = 5, error_type =  "prop", algorithm = "nmf")

```


# Figure 1: experimental design

## Fig 1a: NTCU viability

```{r}

# create df
ntcu_df <- data.frame(Concentration = c("0", "act", "0.1", "0.25", "0.5", "0.75", "1", "2"),
                 mean = c(100, 97.65, 86.52, 58.7, 41.66, 25.13, 24.56, 4.28),
                 sd = c(2.43, 8.17, 8.14, 4.88, 5.44, 10.15, 6.35, 1.01))

ntcu_df$Concentration <- factor(ntcu_df$Concentration, levels=unique(ntcu_df$Concentration))

# plot
fig1a <- ggplot(ntcu_df, aes(x=Concentration, y=mean, group = 1)) + 
  geom_line() + 
  geom_point() +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, color = "darkgray") +
  scale_linetype_manual(name = NULL) +
  scale_y_continuous(limits = c(0, 120), breaks = seq(0, 100, by = 20)) +
  ggtitle("Viability - NTCU day 7") +
  labs(y = "Percent Viability") +
  theme_bw() +
  theme(legend.position = "inside",
      legend.position.inside = c(0.85,0.7))

fig1a

```

## Fig 1b: NTCU LA-QPCR

```{r}

ntcu_df <- data.frame(Treatment = c(rep(c("Control", "Acetone 1%", "NTCU 0.5uM 1h", "NTCU 0.5 uM 24h"), 2)),
                 lm = c(rep("without", 4), rep("with", 4)),
                 mean = c(0.01, -0.01, 0.95, 0.74, -0.10, -0.07, 0.23, 0.08),
                 sd = c(0.02, 0.17, 0.18, 0.20, NA, 0.06, 0.02, 0.08))

ntcu_df$Treatment <- factor(ntcu_df$Treatment, levels=unique(ntcu_df$Treatment))

fig1b <- ggplot(ntcu_df, aes(fill=lm, y=mean, x=Treatment)) + 
  geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, color = "darkgray", position=position_dodge(.9)) +
  scale_fill_manual("liver microsome +\nNADPH status", values = c("with" = "steelblue2", "without" = "red2")) +
  geom_text(aes(x = Treatment, y = mean,label=mean), vjust=-0.25, position=position_dodge(.9), size = 3) +
  ggtitle("Mitochondrial LA-QPCR") +
  labs(y = "lesion / 10 kb") +
  theme_bw() +
  theme(legend.position = "inside",
      legend.position.inside = c(0.25,0.7))

fig1b

```

# Figure 2: SBS patterns for BaP and NTCU treated samples are distinct from controls

## Figure 2a - example mutation profiles

```{r}

fig2a <- plot_sample_counts(musica_all, c("BaP_5nM_DMSO_rep2", "NTCU_0.5nM_Acetone_rep2", "NNK_0.01nM_DMSO_rep2", "NNK_1.0nM_DMSO_rep2", "control_rep2"), "SBS96", same_scale = FALSE, text_size = 12)

fig2a

```

## Figure 2b - correlation plot

```{r}

# extract counts
count_table_for_correlation <- get_count_table(tables(musica_all)$SBS96)

# rename columns
colnames(count_table_for_correlation) <- c("BaP 1", "BaP 2", "BaP 3", "BaP 4", "control acetone 1", "control acetone 2", "control acetone 3", "control DMSO 1", "controlDMSO 2", "control DMSO 3", "control no solvent 1", "control no solvent 2", "control no solvent 3", "NNK low 1", "NNL low 2", "NNK low 3", "NNK low 4", "NNK high 1", "NNK high 2", "NNK high 3", "NNK high 4", "NTCU 1", "NTCU 2", "NTCU 3", "NTCU 4")

# reorder
count_table_for_correlation <- sbs_counts[,c(11,12,13,5,6,7,8,9,10,1,2,3,22,23,24,25,14,15,16,17,18,19,20,21)]

# calculate correlations
correlation <- cosine(count_table_for_correlation)

fig2b <- corrplot(correlation, is.corr = FALSE, type="lower", method = "color", outline = TRUE, order="original",
         tl.col = "black", col=rev(brewer.pal(n=7, name="Spectral")))

fig2b


```

## Figure 2c - total mutational load

```{r}

# extract total mutation counts
sbs_counts <- get_count_table(tables(musica_all)$SBS96)
mut_counts <- colSums(sbs_counts)

# control counts
count_combined_control_df <- data.frame(exposure=c(rep("BaP", 4), rep("control", 3), rep("control", 3), rep("control", 3), rep("NNK_low", 4), rep("NNK_high", 4), rep("NTCU", 4)), num_mutations=mut_counts)
count_combined_control_df$exposure <- as.factor(count_combined_control_df$exposure)

# add log10
count_combined_control_df$log_muts <- log10(count_combined_control_df$num_mutations)

# add log2
count_combined_control_df$log2_muts <- log2(count_combined_control_df$num_mutations)

# specify comparisons
comparisons <- list(c("NNK_low", "NNK_high"), c("control", "NNK_high"), c("control", "NNK_low"), c("control", "NTCU"), c("control", "BaP") )

# compute t test
p_vals <- t_test(count_combined_control_df, log_muts ~ exposure, comparisons = comparisons, p.adjust.method = "BH", detailed = TRUE)
p_vals <- p_vals %>% add_y_position()

p_vals_log2 <- t_test(count_combined_control_df, log2_muts ~ exposure, comparisons = comparisons, p.adjust.method = "BH", detailed = TRUE)

# plot
fig2c <- ggplot(count_combined_control_df, aes(x=exposure, y=num_mutations)) + geom_boxplot(outlier.shape = NA) + scale_y_log10(labels = scales::label_comma()) + geom_jitter(size = 3) + theme_classic() + labs(title="Total SBS Mutations", x="Exposure", y="Number of Mutations") + scale_x_discrete(limits=c("control", "BaP", "NTCU", "NNK_low", "NNK_high"))  + stat_pvalue_manual(p_vals, label = "p.adj") + theme(legend.position="none", text = element_text(size = 17))

fig2c

# calculate fold changes and CIs
raw_fold_change_NTCU <- 1/(exp(-0.561))
raw_CI_high_NTCU <- 1/(exp(-0.706))
raw_CI_low_NTCU <- 1/(exp(-0.416))

raw_fold_change_BaP <- exp(0.838)
raw_CI_low_BaP <- exp(0.296)
raw_CI_high_BaP <- exp(1.38)

```

## Figure 2d - SNR

```{r signal to noise ratio}

# mutation counts for each group
control_counts <- sbs_counts[,c(5:13)]
BaP_counts <- sbs_counts[,c(1:4)]
NTCU_counts <- sbs_counts[,c(22:25)]
NNK_counts <- sbs_counts[,14:21]

# mean mutation counts per group
M_control <- as.data.frame(rowMeans(control_counts))
M_BaP <- as.data.frame(rowMeans(BaP_counts))
M_NTCU <- as.data.frame(rowMeans(NTCU_counts))
M_NNK <- as.data.frame(rowMeans(NNK_counts))

# sd for each group
sd_control <- sd(control_counts)
sd_BaP <- sd(BaP_counts)
sd_NTCU <- sd(NTCU_counts)
sd_NNK <- sd(NNK_counts)

# calculate SNR
SNR_BaP <- norm((M_BaP - M_control), type = "2") / (sd_BaP + sd_control)
SNR_NTCU <- norm((M_NTCU - M_control), type = "2") / (sd_NTCU + sd_control)
SNR_NNK <- norm((M_NNK - M_control), type = "2") / (sd_NNK + sd_control)

# create df
SNR_df <- data.frame(Group = c("BaP", "NTCU", "NNK"), SNR = c(SNR_BaP, SNR_NTCU, SNR_NNK))

# plot
fig2d <- ggplot(data=SNR_df, aes(x=Group, y=SNR)) + 
  geom_bar(stat="identity", color = "black", fill = "black") + 
  scale_x_discrete(limits=c("BaP", "NTCU", "NNK")) +
  scale_y_continuous(limits = c(0,10), breaks = c(0,2,4,6,8,10))+
  geom_text(aes(label=round(SNR, 2)), vjust=-0.7, size=4) +
  theme_classic() +
  theme(text = element_text(size = 17)) +
  labs(title = "Signal to Noise Ratio", x = "Treatment Group", y = "SNR") +
  geom_hline(yintercept = 2, linewidth = 1, col = "red3")

fig2d


```

# Predict SBS signatures

```{r discover signatures}

discover_signatures(musica = musica_all, modality = "SBS96", 
                    num_signatures = 3, algorithm = "nmf", 
                    nstart = 10)

```

```{r reorder signatures}

# change the order in the signatures and the exposures

# rearrange the signatures 

# access the signatures
sig_1 <- signatures(musica_all, "result", "SBS96", "nmf3")[,1]
sig_2 <- signatures(musica_all, "result", "SBS96", "nmf3")[,2]
sig_3 <- signatures(musica_all, "result", "SBS96", "nmf3")[,3]

# put back in new order
signatures(musica_all, "result", "SBS96", "nmf3")[,1] <- sig_3
signatures(musica_all, "result", "SBS96", "nmf3")[,2] <- sig_1
signatures(musica_all, "result", "SBS96", "nmf3")[,3] <- sig_2

# rearrange the exposures 

# access the exposures
exp_1 <- exposures(musica_all, "result", "SBS96", "nmf3")[1,]
exp_2 <- exposures(musica_all, "result", "SBS96", "nmf3")[2,]
exp_3 <- exposures(musica_all, "result", "SBS96", "nmf3")[3,]

# put back in new order
exposures(musica_all, "result", "SBS96", "nmf3")[1,] <- exp_3
exposures(musica_all, "result", "SBS96", "nmf3")[2,] <- exp_1
exposures(musica_all, "result", "SBS96", "nmf3")[3,] <- exp_2

```

# Figure 3: Deconvolution of SBS signatures from treated AALEs

## Figure 3a - signatures

```{r}

# Plot signatures

fig3a <- plot_signatures(musica_all, "nmf3", same_scale = FALSE)

fig3a

```

## Figure 3b - exposures

```{r}

# plot exposures by group

fig3b <- plot_exposures(musica_all, "nmf3", plot_type = "bar", group_by = "annotation", annotation = "Exposure", proportional = TRUE)
fig3b

```

## Figure 3c - comparison to COSMIC

```{r}

# Comparing to external signatures

fig3c <- compare_cosmic_v3(musica_all, "nmf3", sample_type = "genome", modality = "SBS96", threshold = 0.8, result_rename = "Predicted Signatures")
fig3c

```

## Figure 3d - comparison to Gómez-López et al. 2025

```{r}

# load variants
gl_variants <- as.data.frame(read_excel("../data/Murine_Variant_Calls_Gomez-Lopez_2025.xlsx"))

# add column for variant end position
gl_variants$end <- gl_variants$Pos

# select genome
g <- select_genome("mm10")

# create musica
gl_musica <- create_musica_from_variants(gl_variants, g, start_col = "Pos", chromosome_col = "Chrom", sample_col = "sampleID")

# build SBS table
build_standard_table(gl_musica, g, "SBS96")

# discover signatures (1 signature and 2 signatures)
discover_signatures(gl_musica, "SBS96", 1, algorithm = "nmf")
discover_signatures(gl_musica, "SBS96", 2, algorithm = "nmf")

# combine the two results into one matrix for easier plotting
signatures(gl_musica, "result", "SBS96", "nmf2") <- cbind(signatures(gl_musica, "result", "SBS96", "nmf2"), signatures(gl_musica, "result", "SBS96", "nmf1"))
colnames(signatures(gl_musica, "result", "SBS96", "nmf2")) <- c("Component 1 (NTCU Signature)", "Component 2", "All data")

# plot and compare signatures
fig3d <- compare_results(musica = musica_all, model_id = "nmf3", other_model_id = "nmf2", other_musica = gl_musica, threshold = 0.7, result_rename = "Predicted NTCU signature", other_result_rename = "Gómez-López et al. 2025")
fig3d

```

# Supplementary Figure 1: Cytotoxicity and DNA damage response analysis for BaP and NNK

## Fig S1a: BaP viability analysis
```{r}

bap_df <- data.frame(Treatment = c(rep(c("Control", "DMSO 1%", "BaP 0.2uM", "BaP 1uM", "BaP 5uM", "BaP 25uM", "BaP 50uM"), 2)),
                 lm = c(rep("without", 7), rep("with", 7)),
                 mean = c(100, 98, 99, 102, 90, 83, 66, 92, 88, 69, 47, 44, 43, 37),
                 sd = c(0, 0.86, 2.6, 2.75, 4.56, 0, 0, 1.47, 4.1, 1.56, 7.03, 5.11, 3.91, 14.46))

bap_df$Treatment <- factor(bap_df$Treatment, levels=unique(bap_df$Treatment))

figS1a <- ggplot(bap_df, aes(fill=lm, y=mean, x=Treatment)) + 
    geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, color = "darkgray", position=position_dodge(.9)) +
  scale_fill_manual("liver microsome +\nNADPH status", values = c("with" = "steelblue2", "without" = "red2")) +
  geom_text(aes(x = Treatment, y = mean,label=mean), vjust=-0.25, position=position_dodge(.9), size = 3) +
  ggtitle("BaP Viability") +
  labs(y = "Viability") +
  theme_bw() +
  theme(legend.position = "inside",
      legend.position.inside = c(0.16,0.2))

figS1a

```
## Fig S1b: BaP DNA damage response analysis

```{r}

bap_df <- data.frame(Treatment = c(rep(c("Control", "DMSO 1%", "BaP 5uM 1h", "BaP 5uM 24h", "BaP 5uM 48h", "BaP 5uM 72h"), 2)),
                 lm = c(rep("without", 6), rep("with", 6)),
                 mean = c(0.01, 0.08, 0.09, 0.09, -0.05, 0.27, -0.03, -0.04, -0.03, 0.03, 0.02, 0.4),
                 sd = c(0.02, 0.08, 0.1, 0.11, 0, 0.07, 0, 0, 0, 0, 0, 0.06))

bap_df$Treatment <- factor(bap_df$Treatment, levels=unique(bap_df$Treatment))

figS1b <- ggplot(bap_df, aes(fill=lm, y=mean, x=Treatment)) + 
    geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, color = "darkgray", position=position_dodge(.9)) +
  scale_fill_manual("liver microsome +\nNADPH status", values = c("with" = "steelblue2", "without" = "red2")) +
  geom_text(aes(x = Treatment, y = mean,label=mean), vjust=-0.25, position=position_dodge(.9), size = 3) +
  ggtitle("BaP Mitochondrial LA-QPCR") +
  labs(y = "lesion / 10 kb") +
  theme_bw() +
  theme(legend.position = "inside",
      legend.position.inside = c(0.25,0.7))

figS1b

```

## Fig S1c: NNK Viability analysis

```{r}

nnk_df <- data.frame(Treatment = c(rep(c("Control", "DMSO 1%", "NNK 0.01uM", "NNK 0.1uM", "NNK 1uM", "NNK 10uM", "NNK 100uM"), 2)),
                 lm = c(rep("without", 7), rep("with", 7)),
                 mean = c(100, 97.9, 102, 101.9, 104, 100.8, 81.2, 89.5, 86.4, 90.7, 91.8, 93.1, 86, 39.7),
                 sd = c(0, 7.83, 2.11, 2.19, 0.71, 5.94, 1.36, 3.5, 2.29, 2.8, 2.06, 0.35, 3.97, 0.31))

nnk_df$Treatment <- factor(nnk_df$Treatment, levels=unique(nnk_df$Treatment))

figS1c <- ggplot(nnk_df, aes(fill=lm, y=mean, x=Treatment)) + 
    geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, color = "darkgray", position=position_dodge(.9)) +
  scale_fill_manual("liver microsome +\nNADPH status", values = c("with" = "steelblue2", "without" = "red2")) +
  geom_text(aes(x = Treatment, y = mean,label=mean), vjust=-0.25, position=position_dodge(.9), size = 3) +
  ggtitle("NNK Viability") +
  labs(y = "Viability") +
  theme_bw() +
  theme(legend.position = "inside",
      legend.position.inside = c(0.16,0.2))

figS1c

```

## Fig S1d: NNK DNA damage response analysis

```{r}

nnk_df <- data.frame(Treatment = c(rep(c("Control", "DMSO 1%", "NNK 0.01uM", "NNK 0.1uM", "NNK 1uM", "NNK 10uM", "NNK 100uM"), 2)),
                 lm = c(rep("without", 7), rep("with", 7)),
                 mean = c(0.01, 0.11, 0.05, 0.10, 0.01, 0.01, 0.08, -0.02, -0.01, -0.03, 0.04, 0.04, 0.02, 0.06),
                 sd = c(0.01, 0.06, 0.04, 0.15, 0.08, 0.05, 0.04, 0, 0.08, 0.05, 0.04, 0.04, 0.07, 0))

nnk_df$Treatment <- factor(nnk_df$Treatment, levels=unique(nnk_df$Treatment))

figS1d <- ggplot(nnk_df, aes(fill=lm, y=mean, x=Treatment)) + 
    geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2, color = "darkgray", position=position_dodge(.9)) +
  #scale_y_continuous(limits = c(0, 120), breaks = seq(0, 100, by = 20)) +
  scale_fill_manual("liver microsome +\nNADPH status", values = c("with" = "steelblue2", "without" = "red2")) +
  geom_text(aes(x = Treatment, y = mean,label=mean), vjust=-0.25, position=position_dodge(.9), size = 3) +
  ggtitle("NNK Mitochondrial LA-QPCR") +
  labs(y = "lesion / 10 kb") +
  theme_bw() +
  theme(legend.position = "inside",
      legend.position.inside = c(0.8,0.7))

figS1d


```




# Supplementary Figure 2: Analysis of control samples

## Figure S2a - mutational load of controls

```{r}

# plot control sample mutation counts

control_df <- data.frame(Sample=c("acetone1", "acetone2", "acetone3", "DMSO1", "DMSO2", "DMSO3", "control1", "control2", "control3"), num_mutations=mut_counts[5:13], Group = c(rep("acetone", 3), rep("DMSO", 3), rep("control", 3)))

figS2a <- ggplot(data=control_df, aes(x=Sample, y=num_mutations, fill = Group)) + 
  geom_bar(stat="identity") + scale_fill_manual(values=c("#c1272d", "#6495ED", "#eecc16")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ylab("Number of mutations") + scale_x_discrete(limits = c("control1", "control2", "control3", "acetone1", "acetone2", "acetone3", "DMSO1", "DMSO2", "DMSO3"), labels = c(rep("control", 3), rep("acetone", 3), rep("DMSO", 3))) + scale_y_continuous(expand = expansion(mult = c(0, add = 0.05))) + ggtitle("Number of somatic SNVs in controls") + geom_text(aes(label = num_mutations), vjust = -0.5) + theme_classic()  + theme(text = element_text(size = 14)) + theme(legend.position="none")

figS2a

```

## Figure S2b - overlap of somatic SNVs

```{r}

# subset variants to control samples
control_samples <- c("control_rep1", "control_rep2", "control_rep3", "control_DMSO_rep1", "control_DMSO_rep2", "control_DMSO_rep3", "control_acetone_rep1", "control_acetone_rep2", "control_acetone_rep3")
control_variants <- all_variants[all_variants$sample %in% control_samples,]
all_SNVs <- data.frame(sample = control_variants$sample, chrom = control_variants$chr, start = control_variants$start, end = control_variants$end)

# count number of unique SNVs
unique_SNV_counts <- ddply(all_SNVs, .(all_SNVs$chrom, all_SNVs$start, all_SNVs$end), nrow)
names(unique_SNV_counts) <- c("chrom", "start", "end", "Freq")

# now count how many times each frequency occurs
frequency_counts <- ddply(unique_SNV_counts, .(unique_SNV_counts$Freq), nrow)
names(frequency_counts) <- c("num_samples_with_SNV", "Freq")

# plot
figS2b <- ggplot(data=frequency_counts, aes(x=num_samples_with_SNV, y=Freq)) +
  geom_bar(stat="identity", color = "black", fill = "black") + 
  scale_x_discrete(limits=c("1", "2", "3", "4", "5", "6", "7", "8")) +
  geom_text(aes(label=Freq), vjust=-0.3, size=4) +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  labs(title = "Overlap of somatic SNVs in controls", x = "Number of samples with SNV", y = "Frequency")

figS2b

```

## Figure S2c - PCA of germline controls

```{r}

# establish filepaths
C11_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C1-1.vcf.gz"
C12_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C1-2.vcf.gz"
C13_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C1-3.vcf.gz"

C21_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C2-1.vcf.gz"
C22_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C2-2.vcf.gz"
C23_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C2-3.vcf.gz"

C31_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C3-1.vcf.gz"
C32_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C3-2.vcf.gz"
C33_germline_vcf_file_path <- "../data/vcf/haplotype_caller_results/no_gvcf/C3-3.vcf.gz"

# load in variants
C11_germline_variants <- extract_variants_from_vcf_file(vcf_file = C11_germline_vcf_file_path, id = "C1-1", rename = "control_rep1", filter = FALSE)
C12_germline_variants <- extract_variants_from_vcf_file(vcf_file = C12_germline_vcf_file_path, id = "C1-2", rename = "control_rep2", filter = FALSE)
C13_germline_variants <- extract_variants_from_vcf_file(vcf_file = C13_germline_vcf_file_path, id = "C1-3", rename = "control_rep3", filter = FALSE)

C21_germline_variants <- extract_variants_from_vcf_file(vcf_file = C21_germline_vcf_file_path, id = "C2-1", rename = "control_DMSO_rep1", filter = FALSE)
C22_germline_variants <- extract_variants_from_vcf_file(vcf_file = C22_germline_vcf_file_path, id = "C2-2", rename = "control_DMSO_rep2", filter = FALSE)
C23_germline_variants <- extract_variants_from_vcf_file(vcf_file = C23_germline_vcf_file_path, id = "C2-3", rename = "control_DMSO_rep3", filter = FALSE)

C31_germline_variants <- extract_variants_from_vcf_file(vcf_file = C31_germline_vcf_file_path, id = "C3-1", rename = "control_acetone_rep1", filter = FALSE)
C32_germline_variants <- extract_variants_from_vcf_file(vcf_file = C32_germline_vcf_file_path, id = "C3-2", rename = "control_acetone_rep2", filter = FALSE)
C33_germline_variants <- extract_variants_from_vcf_file(vcf_file = C33_germline_vcf_file_path, id = "C3-3", rename = "control_acetone_rep3", filter = FALSE)


# get unique variants in each sample and remove unneeded info
C11 <- unique(data.frame(sample = C11_germline_variants$sample, chrom = C11_germline_variants$chr, start = C11_germline_variants$start, end = C11_germline_variants$end))
C12 <- unique(data.frame(sample = C12_germline_variants$sample, chrom = C12_germline_variants$chr, start = C12_germline_variants$start, end = C12_germline_variants$end))
C13 <- unique(data.frame(sample = C13_germline_variants$sample, chrom = C13_germline_variants$chr, start = C13_germline_variants$start, end = C13_germline_variants$end))

C21 <- unique(data.frame(sample = C21_germline_variants$sample, chrom = C21_germline_variants$chr, start = C21_germline_variants$start, end = C21_germline_variants$end))
C22 <- unique(data.frame(sample = C22_germline_variants$sample, chrom = C22_germline_variants$chr, start = C22_germline_variants$start, end = C22_germline_variants$end))
C23 <- unique(data.frame(sample = C23_germline_variants$sample, chrom = C23_germline_variants$chr, start = C23_germline_variants$start, end = C23_germline_variants$end))

C31 <- unique(data.frame(sample = C31_germline_variants$sample, chrom = C31_germline_variants$chr, start = C31_germline_variants$start, end = C31_germline_variants$end))
C32 <- unique(data.frame(sample = C32_germline_variants$sample, chrom = C32_germline_variants$chr, start = C32_germline_variants$start, end = C32_germline_variants$end))
C33 <- unique(data.frame(sample = C33_germline_variants$sample, chrom = C33_germline_variants$chr, start = C33_germline_variants$start, end = C33_germline_variants$end))

# combine all samples
control_germline_SNPs <- rbind(C11, C12, C13, C21, C22, C23, C31, C32, C33)

# get only unique SNPs
unique_control_germline_SNP_counts <- ddply(control_germline_SNPs, .(control_germline_SNPs$chrom, control_germline_SNPs$start, control_germline_SNPs$end), nrow)
names(unique_control_germline_SNP_counts) <- c("chrom", "start", "end", "Freq")

```


```{r}

# add snp codes (individual identifiers) to individual sample snp lists
C11$snp_code <- paste(C11$chrom, C11$start, C11$end, sep = ".")
C12$snp_code <- paste(C12$chrom, C12$start, C12$end, sep = ".")
C13$snp_code <- paste(C13$chrom, C13$start, C13$end, sep = ".")

C21$snp_code <- paste(C21$chrom, C21$start, C21$end, sep = ".")
C22$snp_code <- paste(C22$chrom, C22$start, C22$end, sep = ".")
C23$snp_code <- paste(C23$chrom, C23$start, C23$end, sep = ".")

C31$snp_code <- paste(C31$chrom, C31$start, C31$end, sep = ".")
C32$snp_code <- paste(C32$chrom, C32$start, C32$end, sep = ".")
C33$snp_code <- paste(C33$chrom, C33$start, C33$end, sep = ".")

#determine snps that are in all samples
common_control_germline_snps <- unique_control_germline_SNP_counts[unique_control_germline_SNP_counts$Freq == 9, c(1,2,3)]
common_control_germline_snps <- paste(common_control_germline_snps$chrom, common_control_germline_snps$start, common_control_germline_snps$end, sep = ".")

# remove snps present in all samples from individual sample snp lists
C11_filtered <- C11[!(C11$snp_code %in% common_control_germline_snps),]
C12_filtered <- C12[!(C12$snp_code %in% common_control_germline_snps),]
C13_filtered <- C13[!(C13$snp_code %in% common_control_germline_snps),]

C21_filtered <- C21[!(C21$snp_code %in% common_control_germline_snps),]
C22_filtered <- C22[!(C22$snp_code %in% common_control_germline_snps),]
C23_filtered <- C23[!(C23$snp_code %in% common_control_germline_snps),]

C31_filtered <- C31[!(C31$snp_code %in% common_control_germline_snps),]
C32_filtered <- C32[!(C32$snp_code %in% common_control_germline_snps),]
C33_filtered <- C33[!(C33$snp_code %in% common_control_germline_snps),]

#determine snps that are in only one sample
unique_snps <- unique_control_germline_SNP_counts[unique_control_germline_SNP_counts$Freq == 1, c(1,2,3)]
unique_snps <- paste(unique_snps$chrom, unique_snps$start, unique_snps$end, sep = ".")

# remove snps that are in only one sample from individual sample snp lists
C11_filtered2 <- C11_filtered[!(C11_filtered$snp_code %in% unique_snps),]
C12_filtered2 <- C12_filtered[!(C12_filtered$snp_code %in% unique_snps),]
C13_filtered2 <- C13_filtered[!(C13_filtered$snp_code %in% unique_snps),]

C21_filtered2 <- C21_filtered[!(C21_filtered$snp_code %in% unique_snps),]
C22_filtered2 <- C22_filtered[!(C22_filtered$snp_code %in% unique_snps),]
C23_filtered2 <- C23_filtered[!(C23_filtered$snp_code %in% unique_snps),]

C31_filtered2 <- C31_filtered[!(C31_filtered$snp_code %in% unique_snps),]
C32_filtered2 <- C32_filtered[!(C32_filtered$snp_code %in% unique_snps),]
C33_filtered2 <- C33_filtered[!(C33_filtered$snp_code %in% unique_snps),]

# get list of unique snp codes remaining
snp_list <- c(C11_filtered2$snp_code, C12_filtered2$snp_code, C13_filtered2$snp_code, C21_filtered2$snp_code, C22_filtered2$snp_code, C23_filtered2$snp_code,C31_filtered2$snp_code, C32_filtered2$snp_code, C33_filtered2$snp_code)

# ensure uniqueness and get reference order
snp_list <- unique(snp_list)

# make dataframe of presence of all snps

#C11
C11_present_snps <- data.frame(snp = C11_filtered2$snp_code, presence = rep(1, length(C11_filtered2$snp_code)))
snps_not_in_C11 <- snp_list[!(snp_list %in% C11_present_snps$snp)]
C11_missing_snps_df <- data.frame(snp = snps_not_in_C11, presence = rep(0, length(snps_not_in_C11)))
C11_snp_presence <- rbind(C11_present_snps, C11_missing_snps_df)
C11_snp_presence <- C11_snp_presence[order(C11_snp_presence$snp),]

#C12
C12_present_snps <- data.frame(snp = C12_filtered2$snp_code, presence = rep(1, length(C12_filtered2$snp_code)))
snps_not_in_C12 <- snp_list[!(snp_list %in% C12_present_snps$snp)]
C12_missing_snps_df <- data.frame(snp = snps_not_in_C12, presence = rep(0, length(snps_not_in_C12)))
C12_snp_presence <- rbind(C12_present_snps, C12_missing_snps_df)
C12_snp_presence <- C12_snp_presence[order(C12_snp_presence$snp),]

#C13
C13_present_snps <- data.frame(snp = C13_filtered2$snp_code, presence = rep(1, length(C13_filtered2$snp_code)))
snps_not_in_C13 <- snp_list[!(snp_list %in% C13_present_snps$snp)]
C13_missing_snps_df <- data.frame(snp = snps_not_in_C13, presence = rep(0, length(snps_not_in_C13)))
C13_snp_presence <- rbind(C13_present_snps, C13_missing_snps_df)
C13_snp_presence <- C13_snp_presence[order(C13_snp_presence$snp),]

#C21
C21_present_snps <- data.frame(snp = C21_filtered2$snp_code, presence = rep(1, length(C21_filtered2$snp_code)))
snps_not_in_C21 <- snp_list[!(snp_list %in% C21_present_snps$snp)]
C21_missing_snps_df <- data.frame(snp = snps_not_in_C21, presence = rep(0, length(snps_not_in_C21)))
C21_snp_presence <- rbind(C21_present_snps, C21_missing_snps_df)
C21_snp_presence <- C21_snp_presence[order(C21_snp_presence$snp),]

#C22
C22_present_snps <- data.frame(snp = C22_filtered2$snp_code, presence = rep(1, length(C22_filtered2$snp_code)))
snps_not_in_C22 <- snp_list[!(snp_list %in% C22_present_snps$snp)]
C22_missing_snps_df <- data.frame(snp = snps_not_in_C22, presence = rep(0, length(snps_not_in_C22)))
C22_snp_presence <- rbind(C22_present_snps, C22_missing_snps_df)
C22_snp_presence <- C22_snp_presence[order(C22_snp_presence$snp),]

#C23
C23_present_snps <- data.frame(snp = C23_filtered2$snp_code, presence = rep(1, length(C23_filtered2$snp_code)))
snps_not_in_C23 <- snp_list[!(snp_list %in% C23_present_snps$snp)]
C23_missing_snps_df <- data.frame(snp = snps_not_in_C23, presence = rep(0, length(snps_not_in_C23)))
C23_snp_presence <- rbind(C23_present_snps, C23_missing_snps_df)
C23_snp_presence <- C23_snp_presence[order(C23_snp_presence$snp),]

#C31
C31_present_snps <- data.frame(snp = C31_filtered2$snp_code, presence = rep(1, length(C31_filtered2$snp_code)))
snps_not_in_C31 <- snp_list[!(snp_list %in% C31_present_snps$snp)]
C31_missing_snps_df <- data.frame(snp = snps_not_in_C31, presence = rep(0, length(snps_not_in_C31)))
C31_snp_presence <- rbind(C31_present_snps, C31_missing_snps_df)
C31_snp_presence <- C31_snp_presence[order(C31_snp_presence$snp),]

#C32
C32_present_snps <- data.frame(snp = C32_filtered2$snp_code, presence = rep(1, length(C32_filtered2$snp_code)))
snps_not_in_C32 <- snp_list[!(snp_list %in% C32_present_snps$snp)]
C32_missing_snps_df <- data.frame(snp = snps_not_in_C32, presence = rep(0, length(snps_not_in_C32)))
C32_snp_presence <- rbind(C32_present_snps, C32_missing_snps_df)
C32_snp_presence <- C32_snp_presence[order(C32_snp_presence$snp),]

#C33
C33_present_snps <- data.frame(snp = C33_filtered2$snp_code, presence = rep(1, length(C33_filtered2$snp_code)))
snps_not_in_C33 <- snp_list[!(snp_list %in% C33_present_snps$snp)]
C33_missing_snps_df <- data.frame(snp = snps_not_in_C33, presence = rep(0, length(snps_not_in_C33)))
C33_snp_presence <- rbind(C33_present_snps, C33_missing_snps_df)
C33_snp_presence <- C33_snp_presence[order(C33_snp_presence$snp),]

# complete matrix of snp presence
master_snp_list <- cbind(C11_snp_presence, C12_snp_presence[,2], C13_snp_presence[,2], C21_snp_presence[,2], C22_snp_presence[,2], C23_snp_presence[,2], C31_snp_presence[,2], C32_snp_presence[,2], C33_snp_presence[,2])

# set row and column names
rownames(master_snp_list) <- master_snp_list$snp
master_snp_list <- master_snp_list[,-c(1)]
colnames(master_snp_list) <- c("C11", "C12", "C13", "C21", "C22", "C23", "C31", "C32", "C33")


```

```{r}

snp.norm = function(x) {
   u <- mean(x)
   p <- u / 2
   (x - u) / sqrt(p * (1 - p))
}

# calculate PCA
snp_list_norm <- apply(master_snp_list, 1, snp.norm)
snp_list_norm <- t(snp_list_norm)
snp_list_norm <- na.omit(snp_list_norm)
colnames(snp_list_norm) <- c("control1", "control2", "control3", "DMSO1", "DMSO2", "DMSO3", "acetone1", "acetone2", "acetone3")

pca_norm = prcomp(t(snp_list_norm), center=FALSE, scale=FALSE)

plotting_coordinates <- as.data.frame(pca_norm$x[,1:2])

plotting_coordinates$Group <- c(rep("control", 3), rep("DMSO", 3), rep("accetone", 3))

# plot PCA
figS2c <- ggplot(plotting_coordinates, aes(x = PC1, y = PC2, col = Group)) + 
  geom_point(size = 2.5) + 
  scale_color_manual(values=c("#c1272d", "#6495ED", "#eecc16")) +
  guides(fill = guide_legend(title = "Group", override.aes = aes(label = ""))) +
  geom_text_repel(aes(label = rownames(plotting_coordinates), size = 15),
                   show.legend = FALSE,
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'black') +
  theme_classic() +
  theme(text = element_text(size = 15)) +
  ggtitle("Parental subclone structure")

figS2c

```

## Fig S2d: overlap of germline SNPs in controls
```{r}

# count how many times each frequency occurs
germline_freq_counts <- ddply(unique_control_germline_SNP_counts, .(unique_control_germline_SNP_counts$Freq), nrow)
names(germline_freq_counts) <- c("num_samples_with_SNP", "Freq")

figS2d <- ggplot(data=germline_freq_counts, aes(x=num_samples_with_SNP, y=Freq)) +
  geom_bar(stat="identity", color = "black", fill = "black") + 
  scale_x_discrete(limits=c("1", "2", "3", "4", "5", "6", "7", "8", "9")) +
  geom_text(aes(label=Freq), vjust=-0.3, size=4) +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  labs(title = "Overlap of germline SNPs in controls", x = "Number of samples with SNP", y = "Frequency")

figS2d

```



# Supplementary Figure 3: SBS mutation pattern of all samples

```{r}

# control samples
figS3a <- plot_sample_counts(musica_all, c("control_rep1", "control_rep2", "control_rep3", "control_DMSO_rep1", "control_DMSO_rep2", "control_DMSO_rep3", "control_acetone_rep1", "control_acetone_rep2", "control_acetone_rep3"), "SBS96", show_x_labels = FALSE)
figS3a

# exposure samples
figS3b <- plot_sample_counts(musica_all, c("BaP_5nM_DMSO_rep1", "BaP_5nM_DMSO_rep2", "BaP_5nM_DMSO_rep3", "BaP_5nM_DMSO_rep4", "NTCU_0.5nM_Acetone_rep1", "NTCU_0.5nM_Acetone_rep2", "NTCU_0.5nM_Acetone_rep3", "NTCU_0.5nM_Acetone_rep4", "NNK_0.01nM_DMSO_rep1", "NNK_0.01nM_DMSO_rep2", "NNK_0.01nM_DMSO_rep3", "NNK_0.01nM_DMSO_rep4", "NNK_1.0nM_DMSO_rep1", "NNK_1.0nM_DMSO_rep2", "NNK_1.0nM_DMSO_rep3", "NNK_1.0nM_DMSO_rep4"), "SBS96", show_x_labels = FALSE)
figS3b

```
# Supplementary Figure 5: DBS signature analysis

## Discover DBS signatures

```{r discover DBS signatures}

discover_signatures(musica = musica_all, modality = "DBS78", 
                              num_signatures = 3, algorithm = "nmf", 
                              nstart = 10)

```

## Fig S5a: DBS signatures

```{r}

figS5a <- plot_signatures(musica_all, "nmf3", modality = "DBS78", same_scale = FALSE)
figS5a

```

## Fig S5b: DBS exposures

```{r}

figS5b <- plot_exposures(musica_all, "nmf3", modality = "DBS78", plot_type = "bar", group_by = "annotation", annotation = "Exposure", proportional = TRUE)
figS5b

```

## Fig S5c: Compare DBS to cosmic

```{r}

figS5c <- compare_cosmic_v3(musica_all, "nmf3", sample_type = "genome", modality = "DBS78", threshold = 0.8)
figS5c

```


# Supplementary Figure 6: IND signature analysis

## Discover IND signatures
```{r discover IND signatures}

# discover signautres
discover_signatures(musica = musica_all, modality = "IND83", 
                    num_signatures = 3, algorithm = "nmf", 
                    nstart = 10)

```

## Fig S6a: IND signatures
```{r}

figS6a <- plot_signatures(musica_all, "nmf3", modality = "IND83", same_scale = FALSE)
figS6a 

```

## Fig S6b: IND exposures
```{r}

figS6b <- plot_exposures(musica_all, "nmf3", modality = "IND83", plot_type = "bar", group_by = "annotation", annotation = "Exposure", proportional = TRUE)
figS6b

```
## Fig S6c: Compare IND to cosmic

```{r}

figS6c<- compare_cosmic_v3(musica_all, model_id = "nmf3", sample_type = "genome", modality = "IND83", threshold = 0.8, result_rename = "Predicted Signatures")
figS6c

```


